---
title: "Storage temperature has a considerably larger impact on spoilage of HTST pasteurized fluid milk as compared to HTST temperature"
author: "T.T. LOTT"
date: "08/19/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###load packages
```{r}
library(tidyverse);library(lmerTest);library(data.table);library(knitr);library(emmeans)
library(patternplot);library(car);library(multcomp);library(dplyr);library(ggpubr);library(stringi)
library(nlsMicrobio);library(minpack.lm);library(AICcmodavg); library(grid); library(Ryacas)
```


###Read in Models

##Buchanan Log10
```{r}
# buchanan (in log10 scale)
buchanan_log10 <- LOG10N ~
  LOG10N0 + #base population
  (t >= lag) * (t <= (lag + (LOG10Nmax - LOG10N0) / mumax)) * #if in growth phase
  mumax * (t - lag) + #log-linear growth (assume positive)
  (t >= lag) * (t > (lag + (LOG10Nmax - LOG10N0) / mumax)) * # if in stationary phase (first condition unnecessary)
  (LOG10Nmax - LOG10N0) #take value of Nmax
```

##Gompertz log10
```{r}
# gompertzm (in log10 scale)
gompertzm_log10 <- LOG10N ~ LOG10N0 + 
  (LOG10Nmax - LOG10N0) * exp(-exp(mumax * exp(1) * (lag - t)/((LOG10Nmax - LOG10N0)) + 1))
```

##Baranyi log10
```{r}
# baranyi (in log10 scale)
# (mumax*log(10))
baranyi_log10 <- LOG10N ~ LOG10Nmax + 
  log10((-1 + exp((mumax*log(10)) * lag) + exp((mumax*log(10)) * t))/(exp((mumax*log(10)) * t) - 1 + exp((mumax*log(10)) * lag) * 10^(LOG10Nmax - LOG10N0)))
```

###read in data
```{r}
##Filter by storage/HTST temp
ESL <- read.csv("ShelfLife_Data_Lott2022a.csv", na.strings = c("", "NA"))
```


###Fit data from storage temperature 3C, HTST temperature 75C to growth models and predict to thresholds
```{r}
##Filter by storage/HTST temp
#3C - 75C
APC3_75 <- ESL %>% 
  filter(Storage_Temp == 3) %>% 
  filter(HTST_Temp == 75) %>% 
  filter(Day <= 42) %>% 
  dplyr::select(APC, Day) %>% 
  filter(!is.na(APC)) %>% 
  mutate(Day = Day*24) %>% 
  rename(t = Day) %>% 
  mutate(APC = replace(APC, APC==0, 2.5)) %>%
  mutate(APC = log10(APC)) %>% 
  rename(LOG10N = APC)

APC3_75 <- APC3_75[c("LOG10N", "t")]

#reorder by column index
APC3_75 <- APC3_75[c(2,1)]

APC3_75 <- APC3_75 %>%
  group_by(t) %>% 
  summarise_at(vars(LOG10N), list(LOG10N = mean)) 


##Select model

# Three candidate growth models
# 1. buchanan (in log10 scale)
# 2. gompertzm (in log10 scale)
# 3. baranyi (in log10 scale)

# Fit model 1: buchanan_log10
# Preview the curve for setting up starting values of the parameters.
preview(buchanan_log10, APC3_75, list(LOG10N0 = 1.68,lag = 400, mumax = 0.004, LOG10Nmax = 4.18))
#Fit the data with the model.
fit3_75.buc_LM <- nlsLM(buchanan_log10, APC3_75, trace=T, 
                         list (LOG10N0 = 1.68,lag = 400, mumax = 0.004, LOG10Nmax = 4.18), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))
#did not run

# Fit model 2: gompertzm_log10
preview(gompertzm_log10, APC3_75, list(LOG10N0 = 1.68,lag = 400, mumax = 0.004, LOG10Nmax = 4.18))
fit3_75.gom_LM <- nlsLM(gompertzm_log10, APC3_75, trace=T, 
                           list (LOG10N0 = 1.68,lag = 400, mumax = 0.004, LOG10Nmax = 4.18), 
                           control = nls.control(maxiter = 1000, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 3: baranyi_log10
preview(baranyi_log10, APC3_75, list(LOG10N0 = 1.68,lag = 400, mumax = 0.004, LOG10Nmax = 4.18))
fit3_75.bar_LM <- nlsLM(baranyi_log10, APC3_75, trace=T, 
                         list (LOG10N0 = 1.68,lag = 400, mumax = 0.004, LOG10Nmax = 4.18), 
                         control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Compare across models to select the best-fit model.
candidate_models <- list()
#candidate_models[[1]] <- fit3_90.buc_LM
candidate_models[[1]] <- fit3_75.gom_LM
candidate_models[[2]] <- fit3_75.bar_LM
mod.names <- c("Gompertz", "Barayni")
print(output_bic)

coef(fit3_75.bar_LM)

##Determine time to threshold

my_model <- "Baranyi"

primary_model_data("Baranyi")

my_pars <- list(logN0 = 1.76, logNmax = 6.81, mu = .003, lambda = 307)

my_time <- seq(0, 2000, length = 1000)

static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

static_prediction$simulation

plot(static_prediction) +
  xlab("Storage time (h)") + 
  ylab("Microbial count (log CFU/mL)") +
  theme_gray()

#20,000 CFU/mL
time_to_logcount(static_prediction, 4.301)

#1,000,000 CFU/mL
time_to_logcount(static_prediction, 6.0)
```

###Fit data from storage temperature 3C, HTST temperature 85C to growth models and predict to thresholds
```{r}
##Filter by storage/HTST temp
#3C - 85C
APC3_85 <- ESL %>% 
  filter(Storage_Temp == 3) %>% 
  filter(HTST_Temp == 85) %>% 
  filter(Day <= 42) %>% 
  dplyr::select(APC, Day) %>% 
  filter(!is.na(APC)) %>% 
  mutate(Day = Day*24) %>% 
  rename(t = Day) %>% 
  mutate(APC = replace(APC, APC==0, 2.5)) %>%
  mutate(APC = log10(APC)) %>% 
  rename(LOG10N = APC)

APC3_85 <- APC3_85[c("LOG10N", "t")]

#reorder by column index
APC3_85 <- APC3_85[c(2,1)]

APC3_85 <- APC3_85 %>%
  group_by(t) %>% 
  summarise_at(vars(LOG10N), list(LOG10N = mean)) 

##Select model 
# Three candidate growth models
# 1. buchanan (in log10 scale)
# 2. gompertzm (in log10 scale)
# 3. baranyi (in log10 scale)

# Fit model 1: buchanan_log10
# Preview the curve for setting up starting values of the parameters.
preview(buchanan_log10, APC3_85, list(LOG10N0 = 1.87,lag = 400, mumax = 0.005, LOG10Nmax = 4.61))
# Fit the data with the model.
fit3_85.buc_LM <- nlsLM(buchanan_log10, APC3_85, trace=T, 
                          list (LOG10N0 = 1.87,lag = 400, mumax = 0.005, LOG10Nmax = 4.61), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))
#Did not run

# Fit model 2: gompertzm_log10
preview(gompertzm_log10, APC3_85, list(LOG10N0 = 1.87,lag = 400, mumax = 0.005, LOG10Nmax = 4.61))
fit3_85_C.gom_LM <- nlsLM(gompertzm_log10, APC3_85, trace=T, 
                           list (LOG10N0 = 1.87,lag = 400, mumax = 0.005, LOG10Nmax = 4.61), 
                           control = nls.control(maxiter = 1000, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 3: baranyi_log10
preview(baranyi_log10, APC3_85, list(LOG10N0 = 1.87,lag = 400, mumax = 0.005, LOG10Nmax = 4.61))
fit3_85.bar_LM <- nlsLM(baranyi_log10, APC3_85, trace=T, 
                           list (LOG10N0 = 1.87,lag = 400, mumax = 0.005, LOG10Nmax = 4.61), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))
#Did not run

# Compare across models to select the best-fit model.
candidate_models <- list()
#candidate_models[[1]] <- fit3_85.buc_LM
candidate_models[[1]] <- fit3_85_C.gom_LM
#candidate_models[[2]] <- fit3_75.bar_LM
mod.names <- c("Gompertzm")

print(output_bic)

coef(fit3_85_C.gom_LM)


##determine time to threshold
my_model <- "modGompertz"

primary_model_data("modGompertz")

my_pars <- list(logN0 = 1.75, C = 5.06, mu = 0.004, lambda = 351)

my_time <- seq(0, 2000, length = 1000)

static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

static_prediction$simulation

plot(static_prediction) +
  xlab("Storage time (h)") + 
  ylab("Microbial count (log CFU/mL)") +
  theme_gray()

#20,000 CFU/mL
time_to_logcount(static_prediction, 4.301)

#1,000,000 CFU/mL
time_to_logcount(static_prediction, 6.0)
```

###Fit data from storage temperature 3C, HTST temperature 90C to growth models and predict to thresholds
```{r}
##Filter by storage/HTST temp
#3C - 90C
APC3_90 <- ESL %>% 
  filter(Storage_Temp == 3) %>% 
  filter(HTST_Temp == 90) %>% 
  filter(Day <= 42) %>% 
  dplyr::select(APC, Day) %>% 
  filter(!is.na(APC)) %>% 
  mutate(Day = Day*24) %>% 
  rename(t = Day) %>% 
  mutate(APC = replace(APC, APC==0, 2.5)) %>%
  mutate(APC = log10(APC)) %>% 
  rename(LOG10N = APC)

APC3_90 - APC3_90[c("LOG10N", "t")]

#reorder by column index
APC3_90 <- APC3_90[c(2,1)]

APC3_90 <- APC3_90 %>%
  group_by(t) %>% 
  summarise_at(vars(LOG10N), list(LOG10N = mean)) 

##Select model
# Three candidate growth models
# 1. buchanan (in log10 scale)
# 2. gompertzm (in log10 scale)
# 3. baranyi (in log10 scale)


# Fit model 1: buchanan_log10
# Preview the curve for setting up starting values of the parameters.
preview(buchanan_log10, APC3_90, list(LOG10N0 = 1.41,lag = 400, mumax = 0.005, LOG10Nmax = 3.93))
# Fit the data with the model.
fit3_90.buc_LM <- nlsLM(buchanan_log10, APC3_90, trace=T, 
                          list (LOG10N0 = 1.41,lag = 400, mumax = 0.005, LOG10Nmax = 3.93), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))

# Fit model 2: gompertzm_log10
preview(gompertzm_log10, APC3_90, list(LOG10N0 = 1.41,lag = 400, mumax = 0.005, LOG10Nmax = 4.18))
fit3_90_C.gom_LM <- nlsLM(gompertzm_log10, APC3_90, trace=T, 
                           list (LOG10N0 = 1.41,lag = 400, mumax = 0.005, LOG10Nmax = 4.18), 
                           control = nls.control(maxiter = 1000, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 3: baranyi_log10
preview(baranyi_log10, APC3_90, list(LOG10N0 = 1.41,lag = 400, mumax = 0.005, LOG10Nmax = 4.18))
fit3_90.bar_LM <- nlsLM(baranyi_log10, APC3_90, trace=T, 
                           list (LOG10N0 = 1.41,lag = 400, mumax = 0.005, LOG10Nmax = 4.18), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Compare across models to select the best-fit model.
candidate_models <- list()
candidate_models[[1]] <- fit3_90.buc_LM
candidate_models[[2]] <- fit3_90_C.gom_LM
candidate_models[[3]] <- fit3_90.bar_LM
mod.names <- c("Buchanan", "Gompertzm", "Barayni")
print(output_bic)

coef(fit3_90.buc_LM)

##determine time to thresholds
my_model <- "Trilinear"

primary_model_data("Trilinear")

my_pars <- list(logN0 = 1.51, logNmax = 6.81, mu = 0.004, lambda = 388)

my_time <- seq(0, 2000, length = 1000)

static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

static_prediction$simulation

plot(static_prediction) +
  xlab("Storage time (h)") + 
  ylab("Microbial count (log CFU/mL)") +
  theme_gray()

#20,000 CFU/mL
time_to_logcount(static_prediction, 4.301)

#1,000,000 CFU/mL
time_to_logcount(static_prediction, 6.0)
```



###Fit data from storage temperature 6.5C, HTST temperature 75C to growth models and predict to thresholds
```{r}
##Filter data
APC6.5_75 <- ESL %>% 
  filter(Storage_Temp == 6.5) %>% 
  filter(HTST_Temp == 75) %>% 
  filter(Day != 42) %>% 
  dplyr::select(APC, Day) %>% 
  filter(!is.na(APC)) %>% 
  mutate(Day = Day*24) %>% 
  rename(t = Day) %>% 
  mutate(APC = replace(APC, APC==0, 2.5)) %>%
  mutate(APC = log10(APC)) %>% 
  rename(LOG10N = APC)

APC6.5_75 <- APC6.5_75[c("LOG10N", "t")]

#reorder by column index
APC6.5_75 <- APC6.5_75[c(2,1)]

APC6.5_75 <- APC6.5_75 %>%
  group_by(t) %>% 
  summarise_at(vars(LOG10N), list(LOG10N = mean)) 

##Select model

# Three candidate growth models
# 1. buchanan (in log10 scale)
# 2. gompertzm (in log10 scale)
# 3. baranyi (in log10 scale)

# Fit model 1: buchanan_log10
# Preview the curve for setting up starting values of the parameters.
preview(buchanan_log10, APC6.5_75, list(LOG10N0 = 1.76,lag = 200, mumax = 0.01, LOG10Nmax = 6.28))
# Fit the data with the model.
fit6.5_75.buc_LM <- nlsLM(buchanan_log10, APC6.5_75, trace=T, 
                           list (LOG10N0 = 1.76,lag = 200, mumax = 0.01, LOG10Nmax = 6.28), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 2: gompertzm_log10
preview(gompertzm_log10, APC6.5_75, list(LOG10N0 = 1.76,lag = 200, mumax = 0.01, LOG10Nmax = 6.28))
fit6.5_75_C.gom_LM <- nlsLM(gompertzm_log10, APC6.5_75, trace=T, 
                           list (LOG10N0 = 1.76,lag = 200, mumax = 0.01, LOG10Nmax = 6.28), 
                           control = nls.control(maxiter = 1000, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 3: baranyi_log10
preview(baranyi_log10, APC6.5_75, list(LOG10N0 = 1.75,lag = 200, mumax = 0.01, LOG10Nmax = 6.1))
fit6.5_75.bar_LM <- nlsLM(baranyi_log10, APC6.5_75, trace=T, 
                           list (LOG10N0 = 1.75,lag = 200, mumax = 0.01, LOG10Nmax = 6.1), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Compare across models to select the best-fit model and get parameters
candidate_models <- list()
candidate_models[[1]] <- fit6.5_75.buc_LM
candidate_models[[2]] <- fit6.5_75_C.gom_LM
candidate_models[[3]] <- fit6.5_75.bar_LM
mod.names <- c("Buchanan", "Gompertzm", "Baranyi")
print(output_bic)

coef(fit6.5_75_C.gom_LM)


## Estimate to thresholds with biogrowth package

library(biogrowth)

my_model <- "modGompertz"

primary_model_data("modGompertz")

my_pars <- list(logN0 = 1.81, C = 5, mu = 0.01, lambda = 257)

my_time <- seq(0, 840, length = 1000)

static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

static_prediction$simulation

plot(static_prediction) +
  xlab("Storage time (h)") + 
  ylab("Microbial count (log CFU/mL)") +
  theme_gray()

#20,000 CFU/mL
time_to_logcount(static_prediction, 4.301)

#1,000,000 CFU/mL
time_to_logcount(static_prediction, 6.0)

```

###Fit data from storage temperature 6.5C, HTST temperature 85C to growth models and predict to thresholds
```{r}
##Filter out data

#6.5C - 85C
APC6.5_85 <- ESL %>% 
  filter(Storage_Temp == 6.5) %>% 
  filter(HTST_Temp == 85) %>% 
  filter(Day != 42) %>% 
  dplyr::select(APC, Day) %>% 
  filter(!is.na(APC)) %>% 
  mutate(Day = Day*24) %>% 
  rename(t = Day) %>% 
  mutate(APC = replace(APC, APC==0, 2.5)) %>%
  mutate(APC = log10(APC)) %>% 
  rename(LOG10N = APC)

APC6.5_85 <- APC6.5_85[c("LOG10N", "t")]

#reorder by column index
APC6.5_85 <- APC6.5_85[c(2,1)]

APC6.5_85 <- APC6.5_85 %>%
  group_by(t) %>% 
  summarise_at(vars(LOG10N), list(LOG10N = mean)) 

##Select model

# Three candidate growth models
# 1. buchanan (in log10 scale)
# 2. gompertzm (in log10 scale)
# 3. baranyi (in log10 scale)

# Fit model 1: buchanan_log10
# Preview the curve for setting up starting values of the parameters.
preview(buchanan_log10, APC6.5_85, list(LOG10N0 = 1.65,lag = 200, mumax = 0.01, LOG10Nmax = 7.33))
# Fit the data with the model.
fit6.5_85.buc_LM <- nlsLM(buchanan_log10, APC6.5_85, trace=T, 
                           list (LOG10N0 = 1.65,lag = 200, mumax = 0.01, LOG10Nmax = 7.33), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))

# Fit model 2: gompertzm_log10
preview(gompertzm_log10, APC6.5_85, list(LOG10N0 = 1.65,lag = 200, mumax = 0.01, LOG10Nmax = 7.33))
fit6.5_85_C.gom_LM <- nlsLM(gompertzm_log10, APC6.5_85, trace=T, 
                           list (LOG10N0 = 1.65,lag = 200, mumax = 0.01, LOG10Nmax = 7.33), 
                           control = nls.control(maxiter = 1000, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 3: baranyi_log10
preview(baranyi_log10, APC6.5_85, list(LOG10N0 = 1.65,lag = 200, mumax = 0.01, LOG10Nmax = 7.33))
fit6.5_85.bar_LM <- nlsLM(baranyi_log10, APC6.5_85, trace=T, 
                           list (LOG10N0 = 1.65,lag = 200, mumax = 0.01, LOG10Nmax = 7.33), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


#Compare across models to select the best-fit model
candidate_models <- list()
candidate_models[[1]] <- fit6.5_85.buc_LM
candidate_models[[2]] <- fit6.5_85_C.gom_LM
candidate_models[[3]] <- fit6.5_85.bar_LM

mod.names <- c("Buchanan", "Gompertzm", "Baranyi")
print(output_bic)

coef(fit6.5_85.buc_LM)

##Determine time to threshold 

my_model <- "Trilinear"

primary_model_data("Trilinear")

my_pars <- list(logN0 = 1.65, logNmax = 6.81, mu = 0.01, lambda = 146)

my_time <- seq(0, 840, length = 1000)

static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

static_prediction$simulation

plot(static_prediction) +
  xlab("Storage time (h)") + 
  ylab("Microbial count (log CFU/mL)") +
  theme_gray()

#20,000 CFU/mL
time_to_logcount(static_prediction, 4.301)

#1,000,000 CFU/mL
time_to_logcount(static_prediction, 6.0)
```


###Fit data from storage temperature 6.5C, HTST temperature 90C to growth models and predict to thresholds
```{r}
##Filter by storage/HTST temp
#6.5C - 90C
APC6.5_90 <- ESL %>% 
  filter(Storage_Temp == 6.5) %>% 
  filter(HTST_Temp == 90) %>% 
  filter(Day != 42) %>% 
  dplyr::select(APC, Day) %>% 
  filter(!is.na(APC)) %>% 
  mutate(Day = Day*24) %>% 
  rename(t = Day) %>% 
  mutate(APC = replace(APC, APC==0, 2.5)) %>%
  mutate(APC = log10(APC)) %>% 
  rename(LOG10N = APC)

APC6.5_90 <- APC6.5_90[c("LOG10N", "t")]

#reorder by column index
APC6.5_90 <- APC6.5_90[c(2,1)]

APC6.5_90 <- APC6.5_90 %>%
  group_by(t) %>% 
  summarise_at(vars(LOG10N), list(LOG10N = mean)) 


##Select model

# Three candidate growth models
# 1. buchanan (in log10 scale)
# 2. gompertzm (in log10 scale)
# 3. baranyi (in log10 scale)

# Fit model 1: buchanan_log10
# Preview the curve for setting up starting values of the parameters.
preview(buchanan_log10, APC6.5_90, list(LOG10N0 = 1.33,lag = 200, mumax = 0.01, LOG10Nmax = 5.42))
# Fit the data with the model.
fit6.5_90.buc_LM <- nlsLM(buchanan_log10, APC6.5_90, trace=T, 
                           list (LOG10N0 = 1.33,lag = 200, mumax = 0.01, LOG10Nmax = 5.42), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 2: gompertzm_log10
preview(gompertzm_log10, APC6.5_90, list(LOG10N0 = 1.33,lag = 200, mumax = 0.01, LOG10Nmax = 5.42))
fit6.5_90_C.gom_LM <- nlsLM(gompertzm_log10, APC6.5_90, trace=T, 
                           list (LOG10N0 = 1.33,lag = 200, mumax = 0.01, LOG10Nmax = 5.42), 
                           control = nls.control(maxiter = 1000, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 3: baranyi_log10
preview(baranyi_log10, APC6.5_90, list(LOG10N0 = 1.33,lag = 200, mumax = 0.01, LOG10Nmax = 5.42))
fit6.5_90.bar_LM <- nlsLM(baranyi_log10, APC6.5_90, trace=T, 
                           list (LOG10N0 = 1.33,lag = 200, mumax = 0.01, LOG10Nmax = 5.42), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Compare across models to select the best-fit model.
candidate_models <- list()
candidate_models[[1]] <- fit6.5_90.buc_LM
candidate_models[[2]] <- fit6.5_90_C.gom_LM
candidate_models[[3]] <- fit6.5_90.bar_LM
mod.names <- c("Buchanan", "Gompertzm", "Baranyi")
print(output_bic)

coef(fit6.5_90.bar_LM)

##Determine time to thresholds

my_model <- "Baranyi"

primary_model_data("Baranyi")

my_pars <- list(logN0 = 1.33, logNmax = 6.81, mu = .01, lambda = 136)

my_time <- seq(0, 2000, length = 1000)

static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

static_prediction$simulation

plot(static_prediction) +
  xlab("Storage time (h)") + 
  ylab("Microbial count (log CFU/mL)") +
  theme_gray()

#20,000
time_to_logcount(static_prediction, 4.301)

#1,000,000
time_to_logcount(static_prediction, 6.0)

```



###Fit data from storage temperature 10C, HTST temperature 75C to growth models and predict to thresholds
```{r}
##Filter by storage/HTST temp
#10C - 75C
APC10_75 <- ESL %>% 
  filter(Storage_Temp == 10) %>% 
  filter(HTST_Temp == 75) %>% 
  filter(Day <= 15) %>% 
  dplyr::select(APC, Day) %>% 
  filter(!is.na(APC)) %>% 
  mutate(Day = Day*24) %>% 
  rename(t = Day) %>% 
  mutate(APC = replace(APC, APC==0, 2.5)) %>%
  mutate(APC = log10(APC)) %>% 
  rename(LOG10N = APC)

APC10_75 - APC10_75[c("LOG10N", "t")]

#reorder by column index
APC10_75 <- APC10_75[c(2,1)]

APC10_75 <- APC10_75 %>%
  group_by(t) %>% 
  summarise_at(vars(LOG10N), list(LOG10N = mean)) 

##Select model
# Three candidate growth models
# 1. buchanan (in log10 scale)
# 2. gompertzm (in log10 scale)
# 3. baranyi (in log10 scale)

# Fit model 1: buchanan_log10
# Preview the curve for setting up starting values of the parameters.
preview(buchanan_log10, APC10_75, list(LOG10N0 = 1.76,lag = 100, mumax = 0.03, LOG10Nmax = 7.15))
# Fit the data with the model.
fit10_75.buc_LM <- nlsLM(buchanan_log10, APC10_75, trace=T, 
                          list (LOG10N0 = 1.76,lag = 100, mumax = 0.03, LOG10Nmax = 7.15), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))

# Fit model 2: gompertzm_log10
preview(gompertzm_log10, APC10_75, list(LOG10N0 = 1.76,lag = 100, mumax = 0.03, LOG10Nmax = 7.15))
fit10_75_C.gom_LM <- nlsLM(gompertzm_log10, APC10_75, trace=T, 
                           list (LOG10N0 = 1.76,lag = 100, mumax = 0.03, LOG10Nmax = 7.15), 
                           control = nls.control(maxiter = 1000, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 3: baranyi_log10
preview(baranyi_log10, APC10_75, list(LOG10N0 = 1.76,lag = 100, mumax = 0.03, LOG10Nmax = 7.15))
fit10_75.bar_LM <- nlsLM(baranyi_log10, APC10_75, trace=T, 
                           list (LOG10N0 = 1.76,lag = 100, mumax = 0.03, LOG10Nmax = 7.15), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Compare across models to select the best-fit model.
candidate_models <- list()
candidate_models[[1]] <- fit10_75.buc_LM
candidate_models[[2]] <- fit10_75_C.gom_LM
candidate_models[[3]] <- fit10_75.bar_LM
mod.names <- c("Buchannan", "Gompertzm", "Barayni")
print(output_bic)

coef(fit10_75.buc_LM)

##Time to threshold
my_model <- "Trilinear"

primary_model_data("Trilinear")

my_pars <- list(logN0 = 1.82, logNmax = 7.03, mu = 0.03, lambda = 84)

my_time <- seq(0, 360, length = 1000)

static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

static_prediction$simulation

plot(static_prediction) +
  xlab("Storage time (h)") + 
  ylab("Microbial count (log CFU/mL)") +
  theme_gray()

#20,000 CFU/mL
time_to_logcount(static_prediction, 4.3)

#1,000,000 CFU/mL
time_to_logcount(static_prediction, 6.0)
```

###Fit data from storage temperature 10C, HTST temperature 85C to growth models and predict to thresholds
```{r}
##Filter by storage/HTST temp
#10C - 85C
APC10_85 <- ESL %>% 
  filter(Storage_Temp == 10) %>% 
  filter(HTST_Temp == 85) %>% 
  filter(Day <= 15) %>% 
  dplyr::select(APC, Day) %>% 
  filter(!is.na(APC)) %>% 
  mutate(Day = Day*24) %>% 
  rename(t = Day) %>% 
  mutate(APC = replace(APC, APC==0, 2.5)) %>%
  mutate(APC = log10(APC)) %>% 
  rename(LOG10N = APC)

APC10_85 - APC10_85[c("LOG10N", "t")]

#reorder by column index
APC10_85 <- APC10_85[c(2,1)]

APC10_85 <- APC10_85 %>%
  group_by(t) %>% 
  summarise_at(vars(LOG10N), list(LOG10N = mean)) 

# Three candidate growth models
# 1. buchanan (in log10 scale)
# 2. gompertzm (in log10 scale)
# 3. baranyi (in log10 scale)

# Fit model 1: buchanan_log10
# Preview the curve for setting up starting values of the parameters.
preview(buchanan_log10, APC10_85, list(LOG10N0 = 1.65,lag = 100, mumax = 0.03, LOG10Nmax = 7.16))
# Fit the data with the model.
fit10_85.buc_LM <- nlsLM(buchanan_log10, APC10_85, trace=T, 
                          list (LOG10N0 = 1.65,lag = 100, mumax = 0.03, LOG10Nmax = 7.16), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))

# Fit model 2: gompertzm_log10
preview(gompertzm_log10, APC10_85, list(LOG10N0 = 1.65,lag = 100, mumax = 0.03, LOG10Nmax = 7.16))
fit10_85_C.gom_LM <- nlsLM(gompertzm_log10, APC10_85, trace=T, 
                           list (LOG10N0 = 1.65,lag = 100, mumax = 0.03, LOG10Nmax = 7.16), 
                           control = nls.control(maxiter = 1000, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 3: baranyi_log10
preview(baranyi_log10, APC10_85, list(LOG10N0 = 1.65,lag = 100, mumax = 0.03, LOG10Nmax = 7.16))
fit10_85.bar_LM <- nlsLM(baranyi_log10, APC10_85, trace=T, 
                           list (LOG10N0 = 1.65,lag = 100, mumax = 0.03, LOG10Nmax = 7.16), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Compare across models to select the best-fit model.
candidate_models <- list()
candidate_models[[1]] <- fit10_85.buc_LM
candidate_models[[2]] <- fit10_85_C.gom_LM
candidate_models[[3]] <- fit10_85.bar_LM
mod.names <- c("Buchannan", "Gompertzm", "Barayni")
print(output_bic)

coef(fit10_85_C.gom_LM)

##Determine time to thresholds
my_model <- "modGompertz"

primary_model_data("modGompertz")

my_pars <- list(logN0 = 1.72, C = 5.09, mu = 0.05, lambda = 107)

my_time <- seq(0, 360, length = 1000)

static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

static_prediction$simulation

plot(static_prediction) +
  xlab("Storage time (h)") + 
  ylab("Microbial count (log CFU/mL)") +
  theme_gray()

#20,000 CFU/mL
time_to_logcount(static_prediction, 4.301)

#1,000,000 CFU/mL
time_to_logcount(static_prediction, 6.0)
```

###Fit data from storage temperature 10C, HTST temperature 90C to growth models and predict to thresholds
```{r}
##Filter by storage/HTST temp
#10C - 90C
APC10_90 <- ESL %>% 
  filter(Storage_Temp == 10) %>% 
  filter(HTST_Temp == 90) %>% 
  filter(Day <= 15) %>% 
  dplyr::select(APC, Day) %>% 
  filter(!is.na(APC)) %>% 
  mutate(Day = Day*24) %>% 
  rename(t = Day) %>% 
  mutate(APC = replace(APC, APC==0, 2.5)) %>%
  mutate(APC = log10(APC)) %>% 
  rename(LOG10N = APC)

APC10_90 - APC10_90[c("LOG10N", "t")]

#reorder by column index
APC10_90 <- APC10_90[c(2,1)]

APC10_90 <- APC10_90 %>%
  group_by(t) %>% 
  summarise_at(vars(LOG10N), list(LOG10N = mean)) 

# Three candidate growth models
# 1. buchanan (in log10 scale)
# 2. gompertzm (in log10 scale)
# 3. baranyi (in log10 scale)

# Fit model 1: buchanan_log10
# Preview the curve for setting up starting values of the parameters.
preview(buchanan_log10, APC10_90, list(LOG10N0 = 1.33,lag = 100, mumax = 0.02, LOG10Nmax = 7.54))
# Fit the data with the model.
fit10_90.buc_LM <- nlsLM(buchanan_log10, APC10_90, trace=T, 
                          list (LOG10N0 = 1.33,lag = 100, mumax = 0.02, LOG10Nmax = 7.54), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))
#did not run


# Fit model 2: gompertzm_log10
preview(gompertzm_log10, APC10_90, list(LOG10N0 = 1.33,lag = 100, mumax = 0.02, LOG10Nmax = 7.54))
fit10_90_C.gom_LM <- nlsLM(gompertzm_log10, APC10_90, trace=T, 
                           list (LOG10N0 = 1.33,lag = 100, mumax = 0.02, LOG10Nmax = 7.54), 
                           control = nls.control(maxiter = 1000, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Fit model 3: baranyi_log10
preview(baranyi_log10, APC10_90, list(LOG10N0 = 1.33,lag = 100, mumax = 0.02, LOG10Nmax = 7.54))
fit10_90.bar_LM <- nlsLM(baranyi_log10, APC10_90, trace=T, 
                           list (LOG10N0 = 1.33,lag = 100, mumax = 0.02, LOG10Nmax = 7.54), 
                           control = nls.control(maxiter = 100, minFactor = 1/4096, warnOnly = T),
                           lower = c(LOG10N0 = 0,lag = 0, mumax = 0, LOG10Nmax = 0))


# Compare across models to select the best-fit model.
candidate_models <- list()
#candidate_models[[1]] <- fit10_90.buc_LM
candidate_models[[1]] <- fit10_90_C.gom_LM
candidate_models[[2]] <- fit10_90.bar_LM
mod.names <- c("Gompertzm", "Barayni")
title_string <- paste("Isolate ", "l3-051", "0 ppm", sep=" ")
title_string
output_bic <- bictab(cand.set = candidate_models, modnames = mod.names, sort = TRUE)
print(title_string)
print(output_bic)

coef(fit10_90.bar_LM)

##determine time to thresholds
my_model <- "Baranyi"

primary_model_data("Baranyi")

my_pars <- list(logN0 = 1.25, logNmax = 6.81, mu = 0.02, lambda = 53)

my_time <- seq(0, 360, length = 1000)

static_prediction <- predict_isothermal_growth(my_model, my_time, my_pars)

static_prediction$simulation

plot(static_prediction) +
  xlab("Storage time (h)") + 
  ylab("Microbial count (log CFU/mL)") +
  theme_gray()

#20,000 CFU/mL
time_to_logcount(static_prediction, 4.3)

#1,000,000 CFU/mL
time_to_logcount(static_prediction, 6.0)
```