---
title: "Storage temperature has a considerably larger impact on spoilage of HTST pasteurized fluid milk as compared to HTST temperature"
author: "T.T. LOTT"
date: "08/18/2022"
output: github_document
editor_options: 
  chunk_output_type: inline
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Load Packages
```{r}
library(dplyr)
library(ggplot2)
library(tidyverse)
library(lme4)
library(lmerTest)
library(rstatix)
library(emmeans)
library(car)
library(ggpubr)
library(effectsize)
```

###Read Data
```{r}
##CSV file with raw data
ESLraw <- read.csv("ShelfLife_Data_Lott2022a.csv", na.strings = c("", "NA"))
```


###Initial Data Transformations
```{r}
##Log Transform and set 0s to detection limit (per plate) (20 CFU/mL) * 25% = 5 CFU/mL
ESL <- ESLraw %>% 
  mutate(APC = replace(APC, APC==0, 5)) %>%
  mutate(APC = log10(APC)) %>% 
  mutate(DayInt_SPC = replace(DayInt_SPC, DayInt_SPC==0, 5)) %>%
  mutate(DayInt_SPC = log10(DayInt_SPC))


##set HTS Temp and Trial as factors
ESL$HTST_Temp <- as.factor(ESL$HTST_Temp)

ESL$Trial <- as.factor(ESL$Trial)

##Make combined variables for trial & storage temp and trial, storage temp, & HTST Temp
ESL$PastBatch <- paste(ESL$Trial, ESL$HTST_Temp, sep="-")

ESL$StoreBatch <- paste(ESL$Trial, ESL$HTST_Temp, ESL$Storage_Temp, sep="-")

#Filter out APC Data, Only include days for each storage temperature that were measured for at least 2 trials 
ESLsub <- ESL %>% 
  filter(!is.na(APC)) %>% 
  filter(Storage_Temp == 3 & Day<= 42 | Storage_Temp == 6.5 & Day<= 35 | Storage_Temp == 10 & Day<= 15 & Trial != 1) %>% 
 mutate(Storage_Temp = as.factor(Storage_Temp)) %>% 
 mutate(Day = as.factor(Day)) %>% 
 mutate(HTST_Temp = as.factor(HTST_Temp)) %>% 
  mutate(sample = paste(Trial, HTST_Temp, sep="-"))

APCrawAvg <- ESLsub %>% 
  ungroup() %>% 
  group_by(HTST_Temp, Storage_Temp, Trial, Day) %>% 
  mutate(APCmean = mean(APC)) %>% 
  ungroup() %>% 
  group_by(HTST_Temp, Storage_Temp, Trial) %>% 
  summarise_at(c("APCmean"), max, na.rm = TRUE)

APCrawAvgtrial <- APCrawAvg %>% 
  ungroup() %>% 
  group_by(HTST_Temp, Storage_Temp) %>% 
  summarise_at(c("APCmean"), mean)

#Create APC column where count is relative to Day 0
#Assign 0 to negative values
APCrel <- ESLsub %>%
  group_by(Trial, Day, HTST_Temp, Storage_Temp) %>% 
  summarise_at(c("APC", "DayInt_SPC"), mean, na.rm = TRUE) %>% 
  mutate(APCr = APC - DayInt_SPC) %>% 
  mutate(APCr = replace(APCr, APCr < 0, 0))

#For finding mean APC relative values
APCrelSummary <- APCrel %>% 
  ungroup() %>% 
  group_by(Day, HTST_Temp, Storage_Temp) %>% 
  summarise_at(c("APCr"), mean)
  
  
#Filter out pH Data, Only include days for each storage temperature that were measured for at least 2 trials
#Transform to H+ for data analysis
ESLpH <- ESL %>% 
  filter(!is.na(pH)) %>% 
  filter(Storage_Temp == 3 & Day<= 42 | Storage_Temp == 6.5 & Day<= 35 | Storage_Temp == 10 & Day<= 15 & Trial != 1) %>% 
  mutate(Storage_Temp = as.factor(Storage_Temp)) %>% 
  mutate(Day = as.factor(Day)) %>% 
  mutate(HTST_Temp = as.factor(HTST_Temp)) %>% 
  mutate(pH = 10^(-pH))

#Add column to make combined variable of HTST Temp and Day
ESLpH$HTST_Day <- paste(ESLpH$HTST_Temp, ESLpH$Day, sep="-")

#Filter out particle size Data, Only include days for each storage temperature that were measured for at least 2 trials, log transform
ESLPS <- ESL %>% 
  filter(!is.na(Particle_Size)) %>% 
  filter(Storage_Temp == 3 & Day<= 42 | Storage_Temp == 6.5 & Day<= 35 | Storage_Temp == 10 & Day<= 15 & Trial != 1) %>% 
  mutate(Particle_Size = log10(Particle_Size)) %>% 
  mutate(Storage_Temp = as.factor(Storage_Temp)) %>% 
  mutate(Day = as.factor(Day)) %>% 
  mutate(HTST_Temp = as.factor(HTST_Temp))

#Filter out CVTA data (Atypical (Y/N)), Only include days for each storage temperature that were measured for at least 2 trials, log transform
ESLcvta <- ESL %>% 
  filter(!is.na(CVTA)) %>% 
  filter(Storage_Temp == 3 & Day<= 42 | Storage_Temp == 6.5 & Day<= 35 | Storage_Temp == 10 & Day<= 15 & Trial != 1) %>% 
  mutate(Storage_Temp = as.factor(Storage_Temp)) %>% 
  mutate(Day = as.factor(Day)) %>% 
  mutate(HTST_Temp = as.factor(HTST_Temp))

#Review how many plates were atypical (Y = atypical, N = not atypical)
table(ESLcvta$CVTA)
```


###APC ANOVA for all storage temperatures and Day 7 & 14 bar plot and all day scatter plot
```{r}
##Data subset

#Average Days 13 and 15 for storage temp 10C to get "day 14" count
#remove duplicate rows for 10c day 14
ESLsubDay14 <- APCrel  %>% 
  mutate(Day = replace(Day, Day == 13 & Storage_Temp==10, 14)) %>% 
  mutate(Day = replace(Day, Day == 15 & Storage_Temp==10, 14)) %>% 
  group_by(Trial, Day, HTST_Temp, Storage_Temp) %>% 
  mutate(APCr = replace(APCr, Day== 14 & Storage_Temp==10, mean(APCr))) %>% 
  select(-APC, -DayInt_SPC) %>% 
  distinct()

#Subset Days 7 and 14 for analysis/plot
ESLsubday7_14 <- ESLsubDay14 %>% 
  filter(Day %in% c("7", "14")) %>% 
  group_by(Day, Storage_Temp) %>% 
  mutate(stdv = sd(APCr)) %>% 
  mutate(avg = mean(APCr)) %>% 
  distinct(Storage_Temp, stdv, avg)

##Scatter plot color by HTST Temp, facet by storage temp
#subset data
APC <- ESLraw[c("Trial", "PlateID", "Day", "HTST_Temp", "Storage_Temp", "APC")] %>% 
  na.omit("APC") %>% 
  group_by(Trial, Day, HTST_Temp, Storage_Temp) %>%
  mutate(APC = replace(APC, APC==0, 5)) %>%
  mutate(APC = log10(APC)) %>% 
  summarise(APCavg = mean(APC))

APCsub <- APC %>% 
  filter(Storage_Temp == 3 & Day<= 42 | Storage_Temp == 6.5 & Day<= 35 | Storage_Temp == 10 & Day<= 15)

#scatter plot
StorTemp_names <- c(
  '3'="3°C",
  '6.5'="6.5°C",
  '10'="10°C")

ggplot(APCsub, aes(x = Day, y = APCavg, color = factor(HTST_Temp), fill = factor(HTST_Temp))) +
  geom_jitter() +
  facet_grid(APCsub$Storage_Temp, labeller = as_labeller(StorTemp_names)) +  
  scale_x_continuous(breaks = c(0,7,14,21,28,35,42)) +
  scale_color_manual(values=c("#56B4E9", "#F0E442", "#D55E00")) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
  xlab("Day") + ylab(expression(~~Log[10]~ CFU/mL)) +
  labs(color = "HTST Temperature (°C)", fill = "HTST Temperature (°C)") +
  geom_smooth(se = FALSE) +
  scale_y_continuous(limits = c(0,8.5))

##Plot APC data by storage temperature for days 7 and 14
ggplot(ESLsubday7_14, aes(x = Day, y = avg, group = Storage_Temp)) +
  geom_col(aes(fill = factor(Storage_Temp)), position = "dodge") +
  xlab("Day") + ylab(expression(Net ~~Log[10]~~ Growth ~~(CFU/mL))) +
  labs(fill = "Storage Temperature (°C)") +
  scale_fill_manual(values=c("#0072B2", "#009E73", "#E69F00")) +
  scale_y_continuous(limits = c(-1,7), expand = c(0, 0)) + 
  geom_errorbar(aes(ymin=avg-stdv, ymax=avg+stdv), width=.5,
                 position=position_dodge(.9))

##Check for potential interactions by plotting

#check for interaction between day and HTST Temp
ESLday14HTST <- ESLsubDay14 %>% 
  group_by(Day, HTST_Temp) %>% 
  summarise_at(c("APCr"), mean, na.rm = TRUE) %>% 
  filter(Day %in% c("0", "7", "14"))

ggplot(data=ESLday14HTST, aes(x=Day, y=APCr, group = HTST_Temp)) +
geom_line((aes(color = factor(HTST_Temp))))
#slopes are similar, no interaction

#check for interaction between day and storage Temp
ESLday14Stor <- ESLsubDay14 %>% 
  group_by(Day, Storage_Temp) %>% 
  summarise_at(c("APCr"), mean, na.rm = TRUE) %>% 
  filter(Day %in% c("0", "7", "14"))

ggplot(data=ESLday14Stor, aes(x=Day, y=APCr, group = Storage_Temp)) +
geom_line((aes(color = factor(Storage_Temp))))
#slopes are different, there is an interaction 

#check for interaction between storage and HTST Temp
ESLday14Temps <- ESLsubDay14 %>% 
  filter(Day %in% c("0", "7", "14")) %>% 
  group_by(HTST_Temp, Storage_Temp) %>% 
  summarise_at(c("APCr"), mean, na.rm = TRUE)

ggplot(data=ESLday14Temps, aes(x=Storage_Temp, y=APCr, group = HTST_Temp)) +
geom_line((aes(color = factor(HTST_Temp))))
#slopes are similar, there is no interaction


##Subset data and perform ANOVA 
#Only Include Days 0, 7, 14 which are only days tested for all Storage Temperatures for ANOVA
ESLaovALL <- ESLsubDay14 %>% 
  filter(Day %in% c("0", "7", "14"))

#ANOVA all storage temps
APCaov <- lmer(APCr ~ HTST_Temp + (Storage_Temp*Day) + (1|Trial), data = ESLaovALL)

#ANOVA Summary
anova(APCaov)

summary(APCaov)

emmeans(APCaov, ~ HTST_Temp + Storage_Temp + Day)

emmeans(APCaov,  pairwise ~ Storage_Temp | Day, adjust = "none")
```


###3C APC ANOVA and Boxplot
```{r}
##Subset data for storage temperature 3C
ESL3CAPC <- ESLsub %>% 
  filter(Storage_Temp %in% c("3"))

##plot 3C APC data
ggplot(data=ESL3CAPC, aes(x=Day, y=APC, fill=as.factor(HTST_Temp))) +
geom_boxplot() +
scale_y_continuous(limits = c(0,8.5), expand = c(0, 0)) +
scale_fill_manual(values=c("#0072B2", "#009E73", "#E69F00")) +
ggtitle("3°C") +
theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
xlab("Day") + ylab(expression(APC~~(Log[10]~ CFU/mL))) +
labs(fill = "HTST Temperature (°C)")


##Subset from Day 21-42
ESL3aov <- ESLsubDay14 %>% 
  filter(Storage_Temp %in% c("3") & Day %in% c("21","28","35","42"))

##3C ANOVA
ESL3aovsum <- lmer(APCr ~ HTST_Temp + Day + (1|Trial), data = ESL3aov)

#Summarize ANOVA
anova(ESL3aovsum)
summary(ESL3aovsum)
emmeans(ESL3aovsum, ~ HTST_Temp + Day)

#effect size
eta_squared(ESL3aovsum, partial = FALSE)
```

###6.5C APC ANOVA
```{r}
##Subset 6.5C Storage Temperature
ESL6.5CAPC <- ESLsub %>% 
  filter(Storage_Temp %in% c("6.5"))

##boxplot of 6.5C APC data
ggplot(data=ESL6.5CAPC, aes(x=Day, y=APC, fill=as.factor(HTST_Temp))) +
geom_boxplot() +
scale_y_continuous(limits = c(0,8.5), expand = c(0, 0)) +
scale_fill_manual(values=c("#0072B2", "#009E73", "#E69F00")) +
ggtitle("6.5°C") +
theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
xlab("Day") + ylab(expression(APC~~(Log[10]~ CFU/mL))) +
labs(fill = "HTST Temperature (°C)")

##subset days where increase is observed (view boxplot)
ESL6.5aov <- ESLsubDay14 %>% 
  filter(Storage_Temp %in% c("6.5") & Day %in% c("14", "21","28","35"))

##6.5C APC ANOVA
ESL6.5aovsum <- lmer(APCr ~ HTST_Temp + Day + (1|Trial), data = ESL6.5aov)

#Summarize
anova(ESL6.5aovsum)
summary(ESL6.5aovsum)
emmeans(ESL6.5aovsum,  pairwise ~ HTST_Temp, adjust = "none")

#effect size
eta_squared(ESL6.5aovsum, partial = FALSE)

```

###10C APC ANOVA
```{r}
##Subset 10C Storage Temperature
ESL10CAPC <- ESLsub %>% 
  filter(Storage_Temp %in% c("10"))

##boxplot of 10C APC data
ggplot(data=ESL10CAPC, aes(x=Day, y=APC, fill=as.factor(HTST_Temp))) +
geom_boxplot() +
scale_y_continuous(limits = c(0,8.5), expand = c(0, 0)) +
scale_fill_manual(values=c("#0072B2", "#009E73", "#E69F00")) +
ggtitle("10°C") +
theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
xlab("Day") + ylab(expression(APC~~(Log[10]~ CFU/mL))) +
labs(fill = "HTST Temperature (°C)")


##subset days where increase is observed (view boxplot)
ESL10aov <- ESLsubDay14 %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("5", "7","9","11", "13", "15"))

##Run 10C ANOVA
ESL10aovsum <- lmer(APCr ~ HTST_Temp + Day + (1|Trial), data = ESL10aov)

#Summarize ANOVA
anova(ESL10aovsum)
summary(ESL10aovsum)
emmeans(ESL10aovsum,  pairwise ~ HTST_Temp, adjust = "none")


#effect size
eta_squared(ESL10aovsum)


##Mean of max trial apc
MeanMax10apc <- APCrel %>% 
  ungroup() %>% 
  filter(Storage_Temp == 10) %>% 
  group_by(Trial, HTST_Temp) %>% 
  summarise_at(c("APC"), max, na.rm = TRUE) 

MeanMax10APCr <- APCrel %>% 
  ungroup() %>% 
  filter(Storage_Temp == 10) %>% 
  group_by(Trial, HTST_Temp) %>% 
  summarise_at(c("APCr"), max, na.rm = TRUE)

mean(MeanMax10APCr$APCr)


```



###ANOVA for all Storage Temperatures for particle size (log10 nm)
```{r}

##Subset data
#Take mean of day 13 and 15 particle size for 10C 
ESLPSDay14 <- ESLPS %>% 
  mutate(Day = replace(Day, Day == 13 & Storage_Temp==10, 14)) %>% 
  mutate(Day = replace(Day, Day == 15 & Storage_Temp==10, 14)) 

#Subset data to only Include Days 0, 7, 14  which are only days tested for all Storage Temperatures
AOVPSall <- ESLPSDay14 %>% 
  filter(Day %in% c("0", "7", "14"))

##ANOVA for particle size for all storage temperatures 
AOVPSallSum <- aov(Particle_Size ~ HTST_Temp + (Storage_Temp * Day), data = AOVPSall)

#summarize ANOVA
summary(AOVPSallSum)

##Pairwise comparison to find within group differences
pwcPS <- AOVPSall %>% tukey_hsd(Particle_Size ~ HTST_Temp + (Storage_Temp * Day))
pwcPS
```


###ANOVA and plot for particle size for Storage Temp 3C
```{r}

##Subset 3C Storage Temperature
ESL3CPS <- ESLPS %>% 
  filter(Storage_Temp %in% c("3"))

##boxplot of 3C particle size data
ggplot(data=ESL3CPS, aes(x=Day, y=Particle_Size, fill=as.factor(HTST_Temp))) +
geom_boxplot() +
scale_y_continuous(limits = c(2,3.6)) +
scale_fill_manual(values=c("#56B4E9", "#F0E442", "#D55E00")) +
ggtitle("3°C") +
theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
theme(axis.text = element_text(size = 10)) +
ylab(expression("Effective Diameter"  ~~log[10]~ (nm))) +
  xlab("Day") +
labs(fill = "HTST Temperature (°C)")


##Subset from Day 21-42 (same days used in APC ANOVA)
PS3aov <- ESL3CPS %>% 
  filter(Storage_Temp %in% c("3") & Day %in% c("21","28","35","42"))

##Run ANOVA
PS3aovsum <- lmer(Particle_Size ~ HTST_Temp * Day + Trial + (1|PastBatch) + (1|PastBatch:Day), data = PS3aov)

#summarize
anova(PS3aovsum)
summary(PS3aovsum)
emmeans(PS3aovsum, ~ HTST_Temp + Day)
emmeans(PS3aovsum,  pairwise ~ HTST_Temp |  Day, adjust = "none")
```

###ANOVA and plot for particle size for Storage Temp 6.5C
```{r}
##Subset 6.5C Storage Temperature
ESL6.5CPS <- ESLPS %>% 
  filter(Storage_Temp %in% c("6.5"))

##boxplot of 6.5C particle size data
ggplot(data=ESL6.5CPS, aes(x=Day, y=Particle_Size, fill=as.factor(HTST_Temp))) +
geom_boxplot() +
scale_y_continuous(limits = c(2,3.6)) +
scale_fill_manual(values=c("#56B4E9", "#F0E442", "#D55E00")) +
ggtitle("6.5°C") +
theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
theme(axis.text = element_text(size = 10)) +
ylab(expression("Effective Diameter"  ~~log[10]~ (nm))) +
  xlab("Day") +
labs(fill = "HTST Temperature (°C)")


##Subset from Day 14-35 (same days used in APC anova)
PS6.5aov <- ESL6.5CPS %>% 
  filter(Storage_Temp %in% c("6.5") & Day %in% c("14","21","28","35"))

##Run ANOVA
PS6.5aovsum <- lmer(Particle_Size ~ HTST_Temp * Day + Trial + (1|PastBatch) + (1|PastBatch:Day), data = PS6.5aov)

##Summarize
anova(PS6.5aovsum)
summary(PS6.5aovsum)
emmeans(PS6.5aovsum, ~ HTST_Temp + Day)
emmeans(PS6.5aovsum,  pairwise ~ HTST_Temp |  Day, adjust = "none")
```

###ANOVA and plot for particle size for Storage Temp 10C
```{r}
##Subset 10C Storage Temperature
ESL10CPS <- ESLPS %>% 
  filter(Storage_Temp %in% c("10"))

##boxplot of 10C particle size data
ggplot(data=ESL10CPS, aes(x=Day, y=Particle_Size, fill=as.factor(HTST_Temp))) +
geom_boxplot() +
scale_y_continuous(limits = c(2,3.6)) +
scale_fill_manual(values=c("#56B4E9", "#F0E442", "#D55E00")) +
ggtitle("10°C") +
theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
theme(axis.text = element_text(size = 10)) +
ylab(expression("Effective Diameter"  ~~log[10]~ (nm))) +
  xlab("Day") +
labs(fill = "HTST Temperature (°C)")

##Subset from Day 5-15
PS10aov <- ESL10CPS %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("5", "7","9","11", "13", "15"))

##Run 10C ANOVA
PS10aovsum <- lmer(Particle_Size ~ HTST_Temp * Day + Trial + (1|PastBatch) + (1|PastBatch:Day), data = PS10aov)

#Summarize ANOVA
anova(PS10aovsum)
summary(PS10aovsum)
emmeans(PS10aovsum, ~ HTST_Temp + Day)
emmeans(PS10aovsum,  pairwise ~ HTST_Temp |  Day, adjust = "none")
```


###ANOVA for pH for all storage temperatures
```{r}

##Average Days 13 and 15 for storage temp 10C to get "day 14" pH
#remove duplicate rows for 10c day 14
ESLsubDay14pH <- ESLpH  %>% 
  mutate(Day = replace(Day, Day == 13 & Storage_Temp==10, 14)) %>% 
  mutate(Day = replace(Day, Day == 15 & Storage_Temp==10, 14)) %>% 
  group_by(Trial, Day, HTST_Temp, Storage_Temp) %>% 
  mutate(pH = replace(pH, Day== 14 & Storage_Temp==10, mean(pH)))


#Only Include Days 0,7, 14 which are only days tested for all Storage Temperatures
pHaovALL <- ESLsubDay14pH %>% 
  filter(Day %in% c("0", "7", "14"))


##ANOVA all storage temps
pHaovALL <- lmer(pH ~ HTST_Temp + (Storage_Temp*Day) + (1|Trial), data = pHaovALL)

#Summarize ANOVA
anova(pHaovALL)

summary(pHaovALL)

emmeans(pHaovALL, ~ HTST_Temp + Storage_Temp + Day)

emmeans(pHaovALL,  pairwise ~ Storage_Temp | Day, adjust = "none")

```



###Kruskal-Wallis and plot for pH for storage temperature 3C
```{r}
##subset 3C data
ESL3CpH <- ESLpH %>% 
  filter(Storage_Temp %in% c("3"))

##boxplot of 3C pH data
ggplot(data=ESL3CpH, aes(x=Day, y=pH, fill=as.factor(HTST_Temp))) +
geom_boxplot() +
scale_y_continuous(limits = c(5.5,7)) +
scale_fill_manual(values=c("#56B4E9", "#F0E442", "#D55E00")) +
ggtitle("3°C") +
theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
theme(axis.text = element_text(size = 10)) +
ylab(expression("pH")) +
  xlab("Day") +
labs(fill = "HTST Temperature (°C)")


##3C kruskal wallace
#Kruskal Wallace and subsequent Dunn's test at day 42
pH3D42KW <- ESL3CpH %>% 
  filter(Storage_Temp %in% c("3") & Day %in% c("42"))

res.kruskal <- pH3D42KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH3D42KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc3D42 <- pH3D42KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc3D42

#Kruskal Wallace and subsequent Dunn's test at day 35
pH3D35KW <- ESL3CpH %>% 
  filter(Storage_Temp %in% c("3") & Day %in% c("35"))

res.kruskal <- pH3D35KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH3D35KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc3D35 <- pH3D35KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc3D35

#Kruskal Wallace and subsequent Dunn's test at day 28
pH3D28KW <- ESL3CpH %>% 
  filter(Storage_Temp %in% c("3") & Day %in% c("28"))

res.kruskal <- pH3D28KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH3D28KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc3D28 <- pH3D28KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc3D28

ph3d28avg <- pH3D28KW %>% 
  ungroup() %>% 
  group_by(HTST_Temp) %>% 
  mutate(pH = -log10(pH)) %>% 
  summarise(pHavg = mean(pH))

#Kruskal Wallace and subsequent Dunn's test at day 21
pH3D21KW <- ESL3CpH %>% 
  filter(Storage_Temp %in% c("3") & Day %in% c("21"))

res.kruskal <- pH3D21KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH3D21KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc3D21 <- pH3D21KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc3D21

#Kruskal Wallace and subsequent Dunn's test at day 14
pH3D14KW <- ESL3CpH %>% 
  filter(Storage_Temp %in% c("3") & Day %in% c("14"))

res.kruskal <- pH3D14KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH3D14KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc3D14 <- pH3D14KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc3D14

#Kruskal Wallace and subsequent Dunn's test at day 7
pH3D7KW <- ESL3CpH %>% 
  filter(Storage_Temp %in% c("3") & Day %in% c("7"))

res.kruskal <- pH3D7KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH3D7KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc3D7 <- pH3D7KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc3D7
```


###Kruskal-Wallis and plot for pH for storage temperature 6.5C
```{r}

##susbset data
ESL6.5CpH <- ESLpH %>% 
  filter(Storage_Temp %in% c("6.5") & Day %in% c("7","14","21","28","35","42")) 
  
ESL6.5CpHplot <- ESLpH %>% 
  mutate(pH = -log10(pH)) %>% 
  filter(Storage_Temp %in% c("6.5") & Day %in% c("0", "7", "14", "21", "28", "35"))


##boxplot of 6.5C pH data
ggplot(data=ESL6.5CpHplot, aes(x=Day, y=pH, fill=as.factor(HTST_Temp))) +
geom_boxplot() +
scale_y_continuous(limits = c(5.5,7)) +
scale_fill_manual(values=c("#56B4E9", "#F0E442", "#D55E00")) +
ggtitle("6.5°C") +
theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
theme(axis.text = element_text(size = 10)) +
ylab(expression("pH")) +
  xlab("Day") +
labs(fill = "HTST Temperature (°C)")


#Kruskal wallace and subsequent Dunn's test for pH for all 6.5C days
pH6.5KW <- ESL6.5CpH %>% 
  filter(Storage_Temp %in% c("6.5"))

res.kruskal <- pH6.5KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH6.5KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc6.5 <- pH6.5KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni") 
pwc6.5


##Kruskal wallace for 6.5C pH
#Kruskal wallace and subsequent Dunn's test for pH for day 28
pH6.5D28KW <- ESL6.5CpH %>% 
  filter(Storage_Temp %in% c("6.5") & Day %in% c("28"))

res.kruskal <- pH6.5D28KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH6.5D28KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc6.5D28 <- pH6.5D28KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc6.5D28

#Kruskal wallace and subsequent Dunn's test for pH for day 21
pH6.5D21KW <- ESL6.5CpH %>% 
  filter(Storage_Temp %in% c("6.5") & Day %in% c("21"))

res.kruskal <- pH6.5D21KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH6.5D21KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc6.5D21 <- pH6.5D21KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc6.5D21

#Kruskal wallace and subsequent Dunn's test for pH for day 14
pH6.5D14KW <- ESL6.5CpH %>% 
  filter(Storage_Temp %in% c("6.5") & Day %in% c("14"))

meanpH6.5cd14 <- pH6.5D14KW %>% 
  group_by(HTST_Temp) %>% 
  summarise(H = mean(pH)) %>% 
  mutate(pHValue = -log10(H))

ph6.5d14avg <- pH6.5D14KW %>% 
  ungroup() %>% 
  group_by(HTST_Temp) %>% 
  mutate(pH = -log10(pH)) %>% 
  summarise(pHavg = mean(pH))

res.kruskal <- pH6.5D14KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH6.5D14KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc6.5D14 <- pH6.5D14KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc6.5D14

#Kruskal wallace and subsequent dunn's test for day 7
pH6.5D7KW <- ESL6.5CpH %>% 
  filter(Storage_Temp %in% c("6.5") & Day %in% c("7"))

res.kruskal <- pH6.5D7KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH6.5D7KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc6.5D7 <- pH6.5D7KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc6.5D7
```


###Kruskal-Wallis and plot for pH for storage temperature 10C
```{r}
##subset data
ESL10CpH <- ESLpH %>% 
 filter(Storage_Temp %in% c("10") & Day %in% c("3", "5", "7", "9","11","13","15"))

ESL10cphplot <- ESLpH %>% 
  mutate(pH = -log10(pH)) %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("0", "3", "5", "7", "9", "11", "13", "15"))


##boxplot of 10C pH data
ggplot(data=ESL10cphplot, aes(x=Day, y=pH, fill=as.factor(HTST_Temp))) +
geom_boxplot() +
scale_y_continuous(limits = c(5.5,7)) +
scale_fill_manual(values=c("#56B4E9", "#F0E442", "#D55E00")) +
ggtitle("10°C") +
theme(plot.title = element_text(face = "bold", hjust = 0.5)) +
theme(axis.text = element_text(size = 10)) +
ylab(expression("pH")) +
  xlab("Day") +
labs(fill = "HTST Temperature (°C)")


##Kruskal wallace for 10C
#Kruskal wallace and subsequent dunn's test for day 15
pH10D15KW <- ESL10CpH %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("15"))

res.kruskal <- pH10D15KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH10D15KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc10D15 <- pH10D15KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc10D15

#Kruskal wallace and subsequent dunn's test for day 13
pH10D13KW <- ESL10CpH %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("13"))

res.kruskal <- pH10D13KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH10D13KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc10D13 <- pH10D13KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc10D13

#Kruskal wallace and subseuqnet dunn's test for day 11
pH10D11KW <- ESL10CpH %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("11"))

res.kruskal <- pH10D11KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH10D11KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc10D11 <- pH10D11KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc10D11

#kruskal wallace and subsequent dunn's test for day 9
pH10D9KW <- ESL10CpH %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("9"))

res.kruskal <- pH10D9KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH10D9KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc10D9 <- pH10D9KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc10D9

#kruskal wallace and subsequent dunn's test for day 7
pH10D7KW <- ESL10CpH %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("7"))

res.kruskal <- pH10D7KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH10D7KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc10D7 <- pH10D7KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc10D7

#kruskal wallace and subsequent dunn's test for day 5
pH10D5KW <- ESL10CpH %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("5"))

res.kruskal <- pH10D5KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH10D5KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc10D5 <- pH10D5KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc10D5

#kruskal wallace and subsequent dunn's test for day 3
pH10D3KW <- ESL10CpH %>% 
  filter(Storage_Temp %in% c("10") & Day %in% c("3"))

res.kruskal <- pH10D3KW %>% kruskal_test(pH ~ HTST_Day)
res.kruskal

pH10D3KW %>% kruskal_effsize(pH ~ HTST_Day)

pwc10D3 <- pH10D3KW %>% 
  dunn_test(pH ~ HTST_Day, p.adjust.method = "bonferroni")
pwc10D3
```


#Summary of raw milk tests plate counts
```{r}
##convert all values to log10
ESLrawmilk <- ESL %>% 
  mutate(Raw_APC = log10(Raw_APC)) %>%
  mutate(LPC = log10(LPC)) %>% 
  mutate(PI = log10(PI)) %>% 
  #set MSC to 25% of detection limit 1cfu/10ml = 0.1cfu/mL*0.25
  mutate(MSC = replace(MSC, MSC==0, (0.025))) %>% 
  mutate(MSC = log10(MSC)) %>% 
  mutate(PSC_10 = replace(PSC_10, PSC_10==0, (0.025))) %>% 
  mutate(PSC_10 = log10(PSC_10)) %>% 
  mutate(PSC_21 = replace(PSC_21, PSC_21==0, (0.025))) %>% 
  mutate(PSC_21 = log10(PSC_21)) 

#mean/range APC
mean(ESLrawmilk$Raw_APC, na.rm =TRUE)
range(ESLrawmilk$Raw_APC, na.rm =TRUE)

#mean/range LPC
mean(ESLrawmilk$LPC, na.rm =TRUE)
range(ESLrawmilk$LPC, na.rm =TRUE)

#mean/range PI
mean(ESLrawmilk$PI, na.rm =TRUE)
range(ESLrawmilk$PI, na.rm =TRUE) 

#mean/range MSC
mean(ESLrawmilk$MSC, na.rm =TRUE)
range(ESLrawmilk$MSC, na.rm =TRUE)

#mean/range MSC
mean(ESLrawmilk$PSC_10, na.rm =TRUE)
range(ESLrawmilk$PSC_10, na.rm =TRUE)

#mean/range MSC
mean(ESLrawmilk$PSC_21, na.rm =TRUE)
range(ESLrawmilk$PSC_21, na.rm =TRUE)

#mean/range for PSC 21 count above detection limit
Trial2 <- ESLrawmilk %>% 
  filter(Trial == 2)

mean(Trial2$PSC_21, na.rm =TRUE)
range(Trial2$PSC_21, na.rm =TRUE)
```
